---
title: "`r params$title`"
author: "`r params$author`"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    number_sections: true
    df_print: kable
  pdf_document:
    toc: true
    number_sections: true
params:
  species_name:    "Unknown species"
  title:           "Pest Invasion Risk Report"
  author:          "pestRisk"
  n_occ:           0
  clean_report:    NULL
  metrics_df:      NULL
  country_summary: NULL
  suit_map_file:   NULL
  risk_map_file:   NULL
  roc_file:        NULL
  imp_file:        NULL
  model_type:      "rf"
  selected_vars:   NULL
  breaks:          !r c(0.33, 0.66)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center")
```

---

# Executive Summary

This report presents the current invasion risk assessment for
**_`r params$species_name`_** generated with the `pestRisk` R package.

The analysis used **`r params$n_occ` cleaned occurrence records** from GBIF
and bioclimatic variables from WorldClim v2.1. Two species distribution models
(SDM) were trained and evaluated. Habitat suitability was predicted across the
study region and classified into **Low / Medium / High** risk zones at
thresholds of `r params$breaks[1]` and `r params$breaks[2]`.

---

# Data Sources

## GBIF Occurrence Data

| Item | Value |
|------|-------|
| Species | *`r params$species_name`* |
| Records after cleaning | `r params$n_occ` |
| Source | Global Biodiversity Information Facility (GBIF) |
| Filter | Georeferenced, no geospatial flags, human observations |

## WorldClim Climate Data

- **Source**: WorldClim v2.1 (Fick & Hijmans 2017)
- **Variables selected after VIF/Pearson filtering**:
  `r paste(params$selected_vars, collapse = ", ")`

---

# Data Cleaning

```{r cleaning_report}
if (!is.null(params$clean_report)) {
  knitr::kable(params$clean_report,
               caption = "Spatial cleaning steps applied to raw GBIF records",
               col.names = c("Step", "Records removed", "Records remaining"))
}
```

---

# Model Performance

```{r metrics}
if (!is.null(params$metrics_df)) {
  knitr::kable(params$metrics_df,
               caption = "SDM performance metrics on held-out test set",
               digits = 3)
}
```

**Metric definitions:**

- **AUC** – Area Under the ROC Curve (0.5 = random, 1.0 = perfect)
- **Sensitivity** – Proportion of true invasion areas correctly identified
- **Specificity** – Proportion of non-invasion areas correctly identified
- **F1** – Harmonic mean of Precision and Recall

```{r roc, fig.cap = "ROC curves for fitted SDM models"}
if (!is.null(params$roc_file) && file.exists(params$roc_file))
  knitr::include_graphics(params$roc_file)
```

---

# Suitability Map

```{r suit_map, fig.cap = paste("Continuous habitat suitability –", params$species_name), out.width = "100%"}
if (!is.null(params$suit_map_file) && file.exists(params$suit_map_file))
  knitr::include_graphics(params$suit_map_file)
```

---

# Risk Classification Map

```{r risk_map, fig.cap = "Invasion risk classification (Low / Medium / High)", out.width = "100%"}
if (!is.null(params$risk_map_file) && file.exists(params$risk_map_file))
  knitr::include_graphics(params$risk_map_file)
```

| Risk class | Suitability range | Interpretation |
|------------|-------------------|----------------|
| Low | 0 – `r params$breaks[1]` | Unlikely to sustain a population |
| Medium | `r params$breaks[1]` – `r params$breaks[2]` | Marginal conditions; possible in warm years |
| High | `r params$breaks[2]` – 1 | Highly suitable; population likely to establish |

---

# Variable Importance

```{r var_imp, fig.cap = "Climate variable importance for pest distribution", out.width = "90%"}
if (!is.null(params$imp_file) && file.exists(params$imp_file))
  knitr::include_graphics(params$imp_file)
```

---

# Regional Risk Ranking

The following table shows the **top vulnerable countries/regions** ranked by
mean habitat suitability.

```{r country_table}
if (!is.null(params$country_summary)) {
  knitr::kable(
    params$country_summary[, c("risk_rank","region","mean_suitability",
                                "max_suitability","pct_high_risk")],
    caption  = paste("Top", nrow(params$country_summary),
                     "highest-risk regions –", params$species_name),
    col.names = c("Rank","Region","Mean suitability",
                  "Max suitability","% High-risk pixels"),
    digits   = 3
  )
}
```

---

# Ecological Interpretation

*_`r params$species_name`_* is a **highly polyphagous agricultural pest** with
high dispersal capacity. The SDM results indicate that warm, humid zones with
mild winters represent the most favourable conditions for population
establishment and spread.

**Agronomic implications:**

1. **Priority surveillance zones** (High risk) should be integrated into
   national phytosanitary monitoring programmes and early-warning systems.
2. **Border inspection** should be reinforced for plant products originating
   from High-risk regions.
3. **Climate change** is expected to shift and expand the suitable range
   poleward and to higher elevations.

---

# Reproducibility & Methods

## Software

```{r session_info, echo = TRUE}
sessionInfo()
```

## Modelling workflow

1. Occurrence data downloaded from GBIF (`rgbif`) with `hasCoordinate = TRUE`
   and no geospatial flags.
2. Spatial cleaning: NA removal, impossible coordinates, Null Island, marine
   points (geodata world boundaries), geographic outlier detection (Mahalanobis
   distance, 3 SD), spatial thinning (10 km grid).
3. WorldClim v2.1 bioclimatic variables downloaded via `geodata::worldclim_global`.
4. Variable selection: Pearson |r| > 0.75 → remove one of each pair; VIF > 5
   → stepwise elimination.
5. Background points: random sampling from non-NA pixels with optional 50 km
   exclusion buffer from presences.
6. Random Forest (500 trees, balanced 1:1, `randomForest` package).
7. MaxEnt via `maxnet` (reg.mult = 1.5, features: linear + quadratic + hinge).
8. Evaluation: AUC (pROC), Accuracy, Sensitivity, Specificity, F1 on 25%
   stratified test set.
9. Risk classification: suitability ∈ [0, 0.33] = Low; [0.33, 0.66] = Medium;
   [0.66, 1] = High.

---

*Report generated by the `pestRisk` R package | FSDM Fès – `r Sys.Date()`*
