---
title: "Introduction to pestRisk: Modelling Pest Invasion Risk"
author: "Meriem Choukri"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to pestRisk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  warning   = FALSE,
  message   = FALSE
)
```

## Overview

`pestRisk` implements a complete Species Distribution Modelling (SDM) workflow
to predict the current invasion risk of agricultural pest species. The package
covers every step from raw data download to publication-ready risk maps and
automatic reports.

```
GBIF occurrences → clean → WorldClim → VIF selection
    → Random Forest + MaxEnt → evaluate → risk map → report
```

The three target species are:

| Species | Common name | Crop impact |
|---------|-------------|-------------|
| *Spodoptera frugiperda* | Fall armyworm | Maize, sorghum, rice |
| *Tuta absoluta* | Tomato leafminer | Tomato, pepper, potato |
| *Ceratitis capitata* | Mediterranean fruit fly | Citrus, stone fruits |

---

## Step 1 – Download GBIF occurrences

```{r download_gbif, eval = FALSE}
library(pestRisk)

# Download fall armyworm records
occ_raw <- download_gbif(
  species = "Spodoptera frugiperda",
  limit   = 2000,
  year    = "2010,2023"
)

dim(occ_raw)
head(occ_raw[, c("species", "decimalLongitude", "decimalLatitude",
                  "year", "country")])
```

Using the built-in example dataset:

```{r load_example}
library(pestRisk)
spodoptera_occ <- read.csv(
  system.file("extdata", "spodoptera_occ.csv", package = "pestRisk")
)
dim(spodoptera_occ)
```

---

## Step 2 – Clean occurrences

Spatial cleaning removes NA coordinates, impossible values, sea points,
geographic outliers, and applies spatial thinning at 20 km to reduce
sampling bias.

```{r clean}
clean_result <- clean_occurrences(
  occ       = spodoptera_occ,
  thin_dist = 20,    # 20 km thinning
  outlier_sd = 3,
  remove_sea = FALSE  # skipped here since we use simulated data
)

print(clean_result$report)
cat("\nFinal occurrences:", nrow(clean_result$data), "\n")
```

---

## Step 3 – Download WorldClim variables

```{r worldclim, eval = FALSE}
# Download 6 key BIO variables at 10 arcmin for Africa + Mediterranean
africa_ext <- c(-20, 55, -40, 40)

clim <- download_worldclim(
  path   = "~/pestRisk_data",
  res    = "10",
  extent = africa_ext,
  vars   = c(1, 4, 5, 6, 12, 15)   # BIO1, BIO4, BIO5, BIO6, BIO12, BIO15
)
terra::plot(clim, main = names(clim))
```

**Why these 6 variables?**

| Variable | Ecological relevance for pest insects |
|----------|---------------------------------------|
| BIO1 (Annual Mean Temp.) | Determines metabolic rate and development speed |
| BIO4 (Temp. Seasonality) | Controls overwinter survival |
| BIO5 (Max Temp. Hottest Month) | Heat stress upper threshold |
| BIO6 (Min Temp. Coldest Month) | Cold-kill threshold for populations |
| BIO12 (Annual Precip.) | Habitat moisture and host plant quality |
| BIO15 (Precip. Seasonality) | Drought stress and host phenology |

---

## Step 4 – Prepare predictors (VIF + Pearson)

```{r prepare, eval = FALSE}
bg_pts <- generate_background_points(
  clim_stack     = clim,
  occurrences_sf = clean_result$data,
  n              = 5000,
  min_dist_km    = 50
)

prep <- prepare_predictors(
  occurrences_sf = clean_result$data,
  background_sf  = bg_pts,
  clim_stack     = clim,
  cor_threshold  = 0.75,
  vif_threshold  = 5
)

cat("Selected variables:", prep$selected_vars, "\n")
cat("VIF values:\n"); print(prep$vif_final)
```

Using the pre-built modelling dataset:

```{r load_model_data}
spodoptera_model_data <- read.csv(
  system.file("extdata", "spodoptera_model_data.csv", package = "pestRisk")
)
str(spodoptera_model_data)
table(spodoptera_model_data$presence)
```

---

## Step 5a – Train Random Forest

```{r train_rf}
rf_out <- train_rf_model(
  model_data = spodoptera_model_data,
  n_trees    = 200,
  balance    = TRUE,
  seed       = 42
)

print(rf_out$importance)
```

---

## Step 5b – Train MaxEnt

```{r train_maxent}
mx_out <- train_maxent(
  model_data      = spodoptera_model_data,
  reg_mult        = 1.5,
  feature_classes = "lqh",
  seed            = 42
)
```

---

## Step 6 – Evaluate models

```{r evaluate}
eval_out <- evaluate_models(
  rf_out, mx_out,
  model_names = c("Random Forest", "MaxEnt"),
  threshold   = 0.5,
  plot_roc    = TRUE
)

print(eval_out$metrics)
```

```{r roc_plot, fig.cap = "ROC curves for RF and MaxEnt models"}
print(eval_out$roc_plot)
```

**Interpretation:**  
For pest surveillance, **Sensitivity** is more important than Specificity —
a missed invasion zone (false negative) is more costly than unnecessary
monitoring (false positive).

---

## Step 7 – Variable importance plot

```{r var_imp, fig.cap = "Random Forest variable importance"}
plot_variable_importance(rf_out, type = "importance")
```

```{r response_curves, fig.cap = "MaxEnt marginal response curves"}
plot_variable_importance(mx_out, type = "response")
```

---

## Step 8 – Predict and visualise the risk map (real WorldClim data)

This section downloads a **small WorldClim extract** (Morocco + West
Africa, 10 arcmin resolution, ~10 MB) so the full prediction pipeline can
run end-to-end with real climate data.

```{r download_small_clim}
study_ext <- c(-20, 20, 0, 40)   # xmin, xmax, ymin, ymax (Morocco + W. Africa)

clim_small <- download_worldclim(
  path   = "worldclim_data",
  res    = "10",
  extent = study_ext,
  vars   = c(1, 12, 15)          # match rf_out$predictors exactly
)

clim_small
```

```{r predict_map}
risk <- predict_risk_map(rf_out, clim_small, breaks = c(0.33, 0.66))

# Quick base-R preview
terra::plot(risk$suitability, main = "Suitability – Spodoptera frugiperda")
```

```{r occ_for_map}
# Keep only the occurrence points that fall inside the study extent
occ_sub <- spodoptera_occ[
  spodoptera_occ$decimalLongitude >= study_ext[1] &
  spodoptera_occ$decimalLongitude <= study_ext[2] &
  spodoptera_occ$decimalLatitude  >= study_ext[3] &
  spodoptera_occ$decimalLatitude  <= study_ext[4], ]

occ_sf <- sf::st_as_sf(occ_sub,
                       coords = c("decimalLongitude", "decimalLatitude"),
                       crs = 4326)
nrow(occ_sf)
```

```{r suitability_map, fig.cap = "Continuous habitat suitability map"}
plot_risk_map(
  risk,
  type            = "continuous",
  occurrences_sf  = occ_sf,
  species_name    = "Spodoptera frugiperda",
  country_borders = FALSE
)
```

```{r risk_class_map, fig.cap = "Classified invasion risk map (Low / Medium / High)"}
plot_risk_map(
  risk,
  type            = "classified",
  species_name    = "Spodoptera frugiperda",
  country_borders = FALSE
)
```

```{r risk_distribution}
freq_tbl <- terra::freq(risk$risk_class)
freq_tbl$pct <- round(100 * freq_tbl$count / sum(freq_tbl$count), 1)
print(freq_tbl)
```

**Interpretation:** The output above shows the percentage of pixels in
each risk category (Low / Medium / High) for the study extent. Areas
classified as **High** (suitability > 0.66) should be prioritised for
phytosanitary surveillance.

---

## Step 9 – Country risk summary

```{r country_risk}
country_tbl <- summarize_country_risk(risk, top_n = 10)
print(country_tbl[, c("risk_rank", "region", "mean_suitability", "pct_high_risk")])
```

**Interpretation:** Countries with a high `mean_suitability` and high
`pct_high_risk` should be prioritised for fall armyworm monitoring
programmes.

---

## Step 10 – Generate automatic report

```{r report, eval = FALSE}
generate_report(
  species_name    = "Spodoptera frugiperda",
  occurrences_sf  = clean_result$data,
  clean_report    = clean_result$report,
  eval_results    = eval_out,
  risk_map_obj    = risk,
  country_summary = country_tbl,
  model_obj       = rf_out,
  output_file     = "Spodoptera_frugiperda_risk_report.html",
  author          = "Meriem Choukri"
)
```

---

## Step 10 – New features: ensemble model & degree-day risk

### 10a. Ensemble suitability (RF + MaxEnt)

`ensemble_suitability()` combines both models, weighted by their AUC
scores from Step 6.

```{r ensemble}
mx_risk <- predict_risk_map(mx_out, clim_small, breaks = c(0.33, 0.66))

rf_auc <- eval_out$metrics$AUC[eval_out$metrics$Model == "Random Forest"]
mx_auc <- eval_out$metrics$AUC[eval_out$metrics$Model == "MaxEnt"]

ens <- ensemble_suitability(
  rf_risk = risk,
  mx_risk = mx_risk,
  rf_auc  = rf_auc,
  mx_auc  = mx_auc
)

terra::plot(ens$ensemble_suitability,
            main = "Ensemble suitability (AUC-weighted RF + MaxEnt)")
```

### 10b. Thermal degree-day risk (generations per year)

`degree_days_risk()` converts WorldClim monthly mean temperature into an
estimate of how many pest generations can be completed per year — a
direct measure of invasion *intensity*, not just presence/absence.

```{r degree_days}
tavg <- geodata::worldclim_global(var = "tavg", res = 10,
                                   path = "worldclim_data")
tavg_small <- terra::crop(tavg, terra::ext(study_ext))

dd <- degree_days_risk(
  clim_stack      = tavg_small,
  t_base          = 12.5,   # S. frugiperda lower threshold
  n_gen_threshold = 450
)

terra::plot(dd$n_generations,
            main = "Estimated generations/year – S. frugiperda")
```

**Interpretation:** Pixels with **≥ 3 generations/year** represent zones
where chemical control is likely to fail due to rapid pest reproduction —
these areas require integrated pest management (IPM) strategies.

---

## Complete one-liner workflow

```{r complete_workflow, eval = FALSE}
library(pestRisk)

species    <- "Tuta absoluta"
study_ext  <- c(-20, 55, -40, 40)
data_path  <- "~/pestRisk_data"

# 1. Download
occ   <- download_gbif(species, limit = 1500, year = "2010,2023")
clim  <- download_worldclim(path = data_path, res = "10",
                              extent = study_ext)

# 2. Clean
clean <- clean_occurrences(occ)

# 3. Background
bg <- generate_background_points(clim, clean$data, n = 5000)

# 4. Predictors
prep <- prepare_predictors(clean$data, bg, clim)

# 5. Models
rf_model <- train_rf_model(prep$model_data)
mx_model <- train_maxent(prep$model_data)

# 6. Evaluate
eval <- evaluate_models(rf_model, mx_model,
                         model_names = c("RF","MaxEnt"))

# 7. Map
risk <- predict_risk_map(rf_model, clim)

# 8. Summary
ctry <- summarize_country_risk(risk)

# 9. Report
generate_report(species, clean$data, clean$report, eval, risk, ctry,
                rf_model, output_file = "tuta_absoluta_report.html")
```

---

## Session info

```{r session}
sessionInfo()
```
